---
title: 'Lab 04: Lasso and IVs: Institutions and development'
output: html_document
date: '2023-03-03'
---

```{r}
suppressMessages({
library(hdm)
library(AER)
data("AJR")
})
```

The original paper: D. Acemoglu, S. Johnson, J. A. Robinson (2001). "Colonial origins of comparative development: an empirical investigation." American Economic Review, 91, 1369–1401.

# Model

$$\begin{equation}
\log (\text{GDP}_i) = \alpha\cdot \text{PropertyRights}_i+X_i'\beta+U_i.
\end{equation}$$

* `GDP` = GDP per capita.
* Property Rights: Institutions, protection from expropriation (the `Exprop` variable).
* Endogeneity: Simultaneity between the output and institutions.
* IV: Mortality of early settlers (`logMort`).

First stage:

$$\begin{equation}
\text{PropertyRights}_i = \pi_1\cdot \text{logMortality}_i+X_i'\Pi_2+V_i
\end{equation}$$

Potential controls in $X_i$:

* `Latitude`
* `Latitude2` (latitude squared)
* `Africa`
* `Asia`
* `Namer` (North America)
* `Samer` (South America)
* `Neo` (Neo-Europes)

# 2SLS estimation with all controls:

```{r}
TSLS<-ivreg(log(GDP)~Exprop+Latitude+Latitude2+Africa+Asia+Namer+Samer+Neo
            | logMort+Latitude+Latitude2+Africa+Asia+Namer+Samer+Neo,data=AJR)
coeftest(TSLS,vcov. = vcovHC(TSLS,type="HC0"))
```

* We use `vcovHC` to correct for heteroskedasticity.
* Insignificant estimate on the main variable `Exprop`.

First stage:
```{r}
FS<-lm(Exprop~logMort+Latitude+Latitude2+Africa+Asia+Namer+Samer+Neo,data=AJR)
coeftest(FS,vcov. = vcovHC(FS,type="HC0"))
```
* Insignificant IV in the first stage.

# Lasso-based approach

We generate the vector/matrices for the dep. variable, main regressor, IV, potential controls
```{r}
Y<-log(AJR$GDP)
X<-model.matrix(Exprop~Latitude+Latitude2+Africa+Asia+Namer+Samer+Neo,data=AJR)[,-1]
Z<-AJR$logMort
D<-AJR$Exprop
```

Lasso implementation:

* `y` = dep. var
* `d` = the main regressor of interest
* `z` = IVs
* `x` = controls
* `select.X=` for controls selection
* `select.Z=` for IVs selection

```{r}
rTSLS<-rlassoIV(y=Y,x=X,d=D,z=Z,select.X = TRUE,select.Z = FALSE)
summary(rTSLS)
```
* Significant estimates
* Marginal effect: 10% increase in GDP due to property rights


# Step by step with Lasso selection (Many controls, few IVs)


### Step 1: Controls and $D$

```{r}
model.D<-rlasso(D~X)
model.D$index
```

* Only `Neo` is selected

Residuals:
```{r}
Dtilde<-model.D$residuals
```

* The residuals in the `rlasso()` output are post-lasso!
* `rlasso()` has the option for post-Lasso: `post=TRUE` (default) or `post=FALSE`.

```{r}
coef(model.D)
```



### Step 2: Controls and $Y$

```{r}
model.Y<-rlasso(Y~X)
Ytilde<-model.Y$residuals
model.Y$index
```

* Selected controls for $Y$:
  - `Latitude2`
  - `Africa`
  - `Samer`
  - `Neo`
  
### Step 3: Controls and $Z$

```{r}
model.Z<-rlasso(Z~X)
Ztilde<-model.Z$residuals
model.Z$index
```

* Selected controls for $Z$:
- `Africa`
- `Neo`

### Step 4: IV estimation using the residuals

Run without the intercept: all variables are in the residual form:
```{r}
rTSLStilde<-ivreg(Ytilde ~ -1 +Dtilde | Ztilde)
coeftest(rTSLStilde,vcov.=vcovHC(rTSLStilde,type="HC0"))
```

---
